Generalized Bradley-Terry models

### 2025-02-19 inst/jss_hyper3_real.Rnw
sushi example

\end{tikzpicture}
\caption{Partial probability\label{tikzabcde} tree for five competitors
  $a$-$e$ with mutually supporting subsets $\left\lbrace a,b,c\right\rbrace$
  [with support term $\lambda$] and $(de)$ [with support term $\mu$]}
\end{center}
\end{figure}

@@

This probability model could easily be modified to account for
specific circumstances.  The cheering effect be asymmetrical (with $a$
helping $b$ but not vice-versa, for example).  The effect could
operate only on specific ordered pairs.  Or perhaps the effect might
have a finite lifetime: if $a$ places $n^\mathrm{th}$, then the
cheering effect is active only for competitors placing
$(n+r)^\mathrm{th}$ or above.

>>>>

\end{tikzpicture}
\caption{Partial probability\label{tikzabcde} tree for five
  competitors $a$-$e$ with two mutually supporting subsets
  $\left\lbrace a,b,c\right\rbrace$ and $\left\lbrace d,
  e\right\rbrace$ having support $\lambda$, $\mu$ respectively}
\end{center}
\end{figure}

@@

This probability model could easily be modified to account for
specific circumstances.  The cheering effect could be asymmetrical
(with $a$ helping $b$ but not vice-versa, for example).  The effect
might operate only on certain ordered pairs, following a directed
graph (cyclic or acyclic) instead of equivalence classes.  Or perhaps
the effect might have a finite lifetime: if $a$ places
$n^\mathrm{th}$, then the cheering effect is active only for
competitors placing $(n+r)^\mathrm{th}$ or above, for some fixed $r$.
There is no reason that $\lambda$ could not be less than one, which
would indicate that some competitors inhibit, rather than help,
others.

The probability model of Figure~\ref{tikzabcde} is applied, in a
simple one-parameter form, to two examples: Formula 1 motor racing,
and also a preference rating.

\section{Red bus, blue bus}

The red bus-blue bus problem was originally stated by
\cite{chipman1960} and discussed by \cite{davidson1970} and
\cite[p153]{alvo2014}.  was \cite{debreu1960}.  The Holiday problem is
discussed in \cite{tverskey1972}.  A traveller has a choice of going
to work by different transport methods.  Here, we can assume that the
traveller has strong but unknown views on different forms of transport
which might incorporate a random component.  If considering a car or
bus, the colour of the bus is of no importance to him.  However, when
offered the choice between a red bus or a blue bus, the colour becomes
important (having nothing else to guide his decision).  Now, he does
not really care what colour the bus is.  If we ask him to rank his
options, it is highly probable that he will put RB and BB
consecutively (because they are essentially indistinguishable).  Can
we quantify the strength of this effect?  To do this, we define a
bespoke function \code{RB\_BB\_LF()} which returns a \code{hyper3}
log-likelihood function corresponding to repeated observations of our
commuter's reported ranks for the five options:

<<defineredbuslf>>=
`RB_BB_LF` <- function(lambda){
    ec <- c(C=1, T=2, RB=3, BB=3, W=4) 
    h <- c(1, 1, lambda, 1)           
    (
        cheering3(v=c("RB", "BB", "C" , "T", "W"), e=ec, h=h)*3 + 
        cheering3(v=c("BB", "RB", "T" , "C", "W"), e=ec, h=h)*2 + 
        cheering3(v=c("T" , "BB", "RB", "C", "W"), e=ec, h=h)*5 + 
        cheering3(v=c("W" , "BB", "RB", "T", "C"), e=ec, h=h)*4 + 
        cheering3(v=c("C" , "RB", "BB", "W", "T"), e=ec, h=h)*3 + 
        cheering3(v=c("BB", "C" , "RB", "T", "W"), e=ec, h=h)*3
    )
}
@

Above, we see from the function body that he reported $RB\succ BB\succ
C\succ T\succ W$ three times [first row], and so on; perhaps his
ranking depends on the weather or how tired he is on any given day.
Observe that in almost every case he ranks RB and BB consecutively.
Function \code{RB_BB_LF()} takes argument \code{lambda} that
quantifies the perceived similarity between RB and BB in the same way
as function \code{oldfirm()} above.  It is straightforward to find a
profile support function for $\lambda$,
Figure~\ref{profilesupportlambdaRBBB}, and formally maximize the
likelihood:

\begin{figure}[htbp]
\begin{center}
<<plotproflike>>=
o <- function(lambda){maxp(RB_BB_LF(lambda), give=1)$`Log-likelihood`} 
lambda <- exp(seq(from=log(0.9), to=log(47), len=17)) 
L <- sapply(lambda, o)
plot(log(lambda), L-max(L), type="b")
abline(h=c(0, -2))
abline(v=0)
@ 
\caption{A\label{profilesupportlambdaRBBB} profile support function
  for the red bus, blue bus parameter $\lambda$.  Vertical line
  indicates null of $\lambda=1$, horizontal lines show maximum support
  and $-2$}
\end{center}
\end{figure}

<<label=maxlikesim, cache=TRUE>>=
(osup <- optimize(o, c(10, 40), maximum=TRUE))
@

So a likelihood ratio test of the null that $\lambda=1$ would be:

<<label=LRT_RB, cache=TRUE>>=
(suppdiff <- o(osup$maximum) - o(1))
@

Easily satisfying Edwards's two-units-of-support criterion;
alternatively, we could observe that this observation lies in the tail
region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2, df=1,
  lower.tail=FALSE),3)}.

\subsection{Sushi}

We now apply the ideas above to a dataset obtained from the {\tt
  preflib} library.

<<readsushi>>=
a <- as.matrix(read.table("sushi.txt"))
head(a)
@

<<defcalculate_sushi_H}>>=
make_sushi_H <- function(a, strength){
    h <- c(1, strength) # strength of support
    sushi_types <- c("ebi", "anago", "maguro", "ika", "uni", "sake", "tamago", "toro", "tekka", "kappa")
    ec <- # equivalence classes
        c(`ebi`  = 1, `anago` = 2, `maguro` = 1, `ika` = 1, `uni` = 1,
          `sake` = 1, `tamago` = 1, `toro` = 2, `tekka-maki` = 2,
          `kappa-maki` = 1)

    H <- hyper3()

    for(i in seq_len(nrow(a))){
        H <- H + cheering3(sushi_types[a[i,]], e=ec, h=h)
    }
    return(H)
}
@

<<tryH,cache=TRUE>>=
wanted <- sample(seq_len(nrow(a)), 50, replace=TRUE, prob=a[,1])
H1888 <- make_sushi_H(a[wanted,-1], strength=1.888)
H1000 <- make_sushi_H(a[wanted,-1], strength=1.000)
@

<<calclikesushi, cache=TRUE>>=
f <- function(strength){
   H <- make_sushi_H(a[wanted,-1], strength=strength)
   maxp(H,give=1)$`Log-likelihood`
}
s <- seq(from=0.4,to=4,len=5)
L <- sapply(s,f)
@

<<plotsushi>>=
plot(s,L-max(L),type='b')
@

<<justf1>>=
f1 <- f(1)
@

<<diffmaxlike>>=
f1 - max(L)
@
